### 2018-03-16 submission/manuscript.Rmd
remove typos and incorporate Pat's edits

An increasing body of literature suggests that both individual and collections of bacteria are associated with the progression of colorectal cancer. As the number of studies investigating these associations increases and the number of subjects in each study increases, a meta-analysis to identify the associations that are the most predictive of disease progression is warranted. We analyzed previously published 16S rRNA gene sequencing data collected from feces and colon tissue.

We quantified the odds ratios (ORs) for individual bacterial taxa that were associated with an individual having tumors relative to a normal colon. Among the fecal samples, there were no taxa that had significant ORs associated with adenoma and there were `r length(crc_OR_stool$measure)` taxa with significant ORs associated with carcinoma. Similarly, among the tissue samples, there were no taxa that had a significant OR associated with adenoma and there were `r length(crc_unmatched_OR_tissue$measure)` taxa with significant ORs associated with carcinoma. Among the significant ORs, the association between individual taxa and tumor diagnosis was equal or below `r ifelse(max(crc_OR_stool$rr) > max(crc_unmatched_OR_tissue$rr), round(max(crc_OR_stool$rr), digits = 2), round(max(crc_unmatched_OR_tissue$rr), digits = 2))`. Because individual taxa had limited association with tumor diagnosis, we trained Random Forest classification models using only the taxa that had significant ORs, using the entire collection of taxa found in each study, and using operational taxonomic units defined based on a 97% similarity threshold. All training approaches yielded similar classification success as measured using the Area Under the Curve. The ability to correctly classify individuals with adenomas was poor and the ability to classify individuals with carcinomas was considerably better using sequences from fecal or tissue.

@@

***Random Forest classification analysis.***
To classify individuals as having normal colons or tumors, we built Random Forest classification models for each dataset and comparison using taxa with significant ORs (after multiple comparison correction), all taxa, or OTUs.  Because no taxa were identified as having a significant OR associated with adenomas using stool samples or tissue samples, classification models based on OR data were not constructed to classify individuals as having normal colons or adenomas. For all models the value of trees included (i.e. ntree) was set to 500 and the number of variables that were randomly tested (i.e. mtry) was set to the square root of the number of taxa or OTUs within the model. Using the square root of the total number of features as the number of features to test has been found to reliably approximate the optimum value after model tuning [@Breiman2001]. All fecal models were built using a 10-fold cross validation (CV) while tissue models were built using 5-fold CV due to study sample size. One exception to this were the models constructed using data from the Weir study, which was built using a 2-fold CV due to the small number of samples sample. For models constructed based on the taxa that had a significant OR or using all of the taxa, we trained the models using a single study and then tested on the remaining studies with AUCs recorded during both train and testing phases. For the models constructed using OTU data, 100 10-fold CVs were run to generate a range of AUCs that could be reasonably expected to occur. The average AUC from these 100 repeats was reported. The Mean Decrease in Accuracy (MDA), a measure of the importance of each taxon to the overall model, was used to rank the taxa used in each model.

>>>>

An increasing body of literature suggests that both individual and collections of bacteria are associated with the progression of colorectal cancer. As the number of studies investigating these associations increases and the number of subjects in each study increases, a meta-analysis to identify the associations that are the most predictive of disease progression is warranted. We analyzed previously published 16S rRNA gene sequencing data collected from feces and colon tissue. We quantified the odds ratios (ORs) for individual bacterial taxa that were associated with an individual having tumors relative to a normal colon. Among the fecal samples, there were no taxa that had significant ORs associated with adenoma and there were `r length(crc_OR_stool$measure)` taxa with significant ORs associated with carcinoma. Similarly, among the tissue samples, there were no taxa that had a significant OR associated with adenoma and there were `r length(crc_unmatched_OR_tissue$measure)` taxa with significant ORs associated with carcinoma. Among the significant ORs, the association between individual taxa and tumor diagnosis was equal or below `r ifelse(max(crc_OR_stool$rr) > max(crc_unmatched_OR_tissue$rr), round(max(crc_OR_stool$rr), digits = 2), round(max(crc_unmatched_OR_tissue$rr), digits = 2))`. Because individual taxa had limited association with tumor diagnosis, we trained Random Forest classification models using only the taxa that had significant ORs, using the entire collection of taxa found in each study, and using operational taxonomic units defined based on a 97% similarity threshold. All training approaches yielded similar classification success as measured using the Area Under the Curve. The ability to correctly classify individuals with adenomas was poor and the ability to classify individuals with carcinomas was considerably better using sequences from fecal or tissue.

@@

***Random Forest classification analysis.***
To classify individuals as having normal colons or tumors, we built Random Forest classification models for each dataset and comparison using taxa with significant ORs (after multiple comparison correction), all taxa, or OTUs.  Because no taxa were identified as having a significant OR associated with adenomas using stool or tissue samples, classification models based on OR data were not constructed to classify individuals as having normal colons or adenomas. For all models, the value of trees included (i.e. ntree) was set to 500 and the number of variables that were randomly tested (i.e. mtry) was set to the square root of the number of taxa or OTUs within the model. Using the square root of the total number of features as the number of features to test has been found to reliably approximate the optimum value after model tuning [@Breiman2001]. All fecal models were built using a 10-fold cross validation (CV) while tissue models were built using 5-fold CV due to study sample size. One exception to this were the models constructed using data from the Weir study, which was built using a 2-fold CV due to the small number of samples. For models constructed based on the taxa that had a significant OR or using all of the taxa, we trained the models using a single study and then tested on the remaining studies with AUCs recorded during both train and testing phases. For the models constructed using OTU data, 100 10-fold CVs were run to generate a range of AUCs that could be reasonably expected to occur. The average AUC from these 100 repeats was reported. The Mean Decrease in Accuracy (MDA), a measure of the importance of each taxon to the overall model, was used to rank the taxa used in each model.

### 2018-03-16 submission/manuscript.tex
remove typos and incorporate Pat's edits

An increasing body of literature suggests that both individual and
collections of bacteria are associated with the progression of
colorectal cancer. As the number of studies investigating these
associations increases and the number of subjects in each study
increases, a meta-analysis to identify the associations that are the
most predictive of disease progression is warranted. We analyzed
previously published 16S rRNA gene sequencing data collected from feces
and colon tissue.

We quantified the odds ratios (ORs) for individual bacterial taxa that
were associated with an individual having tumors relative to a normal
colon. Among the fecal samples, there were no taxa that had significant
ORs associated with adenoma and there were 8 taxa with significant ORs
associated with carcinoma. Similarly, among the tissue samples, there
were no taxa that had a significant OR associated with adenoma and there
were 3 taxa with significant ORs associated with carcinoma. Among the
significant ORs, the association between individual taxa and tumor
diagnosis was equal or below 7.11. Because individual taxa had limited
association with tumor diagnosis, we trained Random Forest
classification models using only the taxa that had significant ORs,
using the entire collection of taxa found in each study, and using
operational taxonomic units defined based on a 97\% similarity
threshold. All training approaches yielded similar classification
success as measured using the Area Under the Curve. The ability to
correctly classify individuals with adenomas was poor and the ability to
classify individuals with carcinomas was considerably better using
sequences from fecal or tissue.

@@

\textbf{\emph{Random Forest classification analysis.}} To classify
individuals as having normal colons or tumors, we built Random Forest
classification models for each dataset and comparison using taxa with
significant ORs (after multiple comparison correction), all taxa, or
OTUs. Because no taxa were identified as having a significant OR
associated with adenomas using stool samples or tissue samples,
classification models based on OR data were not constructed to classify
individuals as having normal colons or adenomas. For all models the
value of trees included (i.e.~ntree) was set to 500 and the number of
variables that were randomly tested (i.e.~mtry) was set to the square
root of the number of taxa or OTUs within the model. Using the square
root of the total number of features as the number of features to test
has been found to reliably approximate the optimum value after model
tuning (53). All fecal models were built using a 10-fold cross
validation (CV) while tissue models were built using 5-fold CV due to
study sample size. One exception to this were the models constructed
using data from the Weir study, which was built using a 2-fold CV due to
the small number of samples sample. For models constructed based on the
taxa that had a significant OR or using all of the taxa, we trained the
models using a single study and then tested on the remaining studies
with AUCs recorded during both train and testing phases. For the models
constructed using OTU data, 100 10-fold CVs were run to generate a range
of AUCs that could be reasonably expected to occur. The average AUC from
these 100 repeats was reported. The Mean Decrease in Accuracy (MDA), a
measure of the importance of each taxon to the overall model, was used
to rank the taxa used in each model.

\textbf{\emph{Statistical Analysis.}} All statistical analysis after
sequence processing utilized the R (v3.4.3) software package (54). For
OTU richness, evenness, and Shannon diversity analysis, values were
power transformed using the rcompanion (v1.11.1) package (55) and
Z-score normalized using the car (v2.1.6) package (56). Testing for OTU
richness, evenness, and Shannon diversity differences utilized linear
mixed-effect models to correct for study, repeat sampling of individuals
(tissue only), and 16S rRNA gene sequence region used using the lme4
(v1.1.15) package (57). ORs were analyzed using both the epiR (v0.9.93)
and metafor (v2.0.0) packages (58, 59) by assessing how many individuals
with and without disease were above and below the overall median value
within each specific study. OR significance testing utilized the
chi-squared test. Community structure differences were calculated using
the Bray-Curtis dissimilarity index and PERMANOVA was used to test for
tumor-associated differences in structure with the vegan (v2.4.4)
package (60). Random Forest models were built using both the caret
(v6.0.77) and randomForest (v4.6.12) packages (61, 62). All figures were
created using both ggplot2 (v2.2.1.9000) and gridExtra (v2.3) packages
(63, 64).

>>>>

An increasing body of literature suggests that both individual and
collections of bacteria are associated with the progression of
colorectal cancer. As the number of studies investigating these
associations increases and the number of subjects in each study
increases, a meta-analysis to identify the associations that are the
most predictive of disease progression is warranted. We analyzed
previously published 16S rRNA gene sequencing data collected from feces
and colon tissue. We quantified the odds ratios (ORs) for individual
bacterial taxa that were associated with an individual having tumors
relative to a normal colon. Among the fecal samples, there were no taxa
that had significant ORs associated with adenoma and there were 8 taxa
with significant ORs associated with carcinoma. Similarly, among the
tissue samples, there were no taxa that had a significant OR associated
with adenoma and there were 3 taxa with significant ORs associated with
carcinoma. Among the significant ORs, the association between individual
taxa and tumor diagnosis was equal or below 7.11. Because individual
taxa had limited association with tumor diagnosis, we trained Random
Forest classification models using only the taxa that had significant
ORs, using the entire collection of taxa found in each study, and using
operational taxonomic units defined based on a 97\% similarity
threshold. All training approaches yielded similar classification
success as measured using the Area Under the Curve. The ability to
correctly classify individuals with adenomas was poor and the ability to
classify individuals with carcinomas was considerably better using
sequences from fecal or tissue.

@@

\textbf{\emph{Random Forest classification analysis.}} To classify
individuals as having normal colons or tumors, we built Random Forest
classification models for each dataset and comparison using taxa with
significant ORs (after multiple comparison correction), all taxa, or
OTUs. Because no taxa were identified as having a significant OR
associated with adenomas using stool or tissue samples, classification
models based on OR data were not constructed to classify individuals as
having normal colons or adenomas. For all models, the value of trees
included (i.e.~ntree) was set to 500 and the number of variables that
were randomly tested (i.e.~mtry) was set to the square root of the
number of taxa or OTUs within the model. Using the square root of the
total number of features as the number of features to test has been
found to reliably approximate the optimum value after model tuning (53).
All fecal models were built using a 10-fold cross validation (CV) while
tissue models were built using 5-fold CV due to study sample size. One
exception to this were the models constructed using data from the Weir
study, which was built using a 2-fold CV due to the small number of
samples. For models constructed based on the taxa that had a significant
OR or using all of the taxa, we trained the models using a single study
and then tested on the remaining studies with AUCs recorded during both
train and testing phases. For the models constructed using OTU data, 100
10-fold CVs were run to generate a range of AUCs that could be
reasonably expected to occur. The average AUC from these 100 repeats was
reported. The Mean Decrease in Accuracy (MDA), a measure of the
importance of each taxon to the overall model, was used to rank the taxa
used in each model.

\textbf{\emph{Statistical Analysis.}} All statistical analysis after
sequence processing utilized the R (v3.4.3) software package (54). For
OTU richness, evenness, and Shannon diversity analysis, values were
power transformed using the rcompanion (v1.11.1) package (55) and
Z-score normalized using the car (v2.1.6) package (56). Testing for OTU
richness, evenness, and Shannon diversity differences utilized linear
mixed-effect models to correct for study, repeat sampling of individuals
(tissue only), and 16S rRNA gene sequence region used using the lme4
(v1.1.15) package (57). ORs were analyzed using both the epiR (v0.9.93)
and metafor (v2.0.0) packages (58, 59) by assessing how many individuals
with and without disease were above and below the overall median value
within each specific study. OR significance testing utilized the
chi-squared test. Community structure differences were calculated using
the Bray-Curtis dissimilarity index and PERMANOVA was used to test for
tumor-associated differences in structure with the vegan (v2.4.5)
package (60). Random Forest models were built using both the caret
(v6.0.78) and randomForest (v4.6.12) packages (61, 62). All figures were
created using both ggplot2 (v2.2.1) and gridExtra (v2.3) packages (63,
64).
